---
title: "Making Plots Interactive"
---

::: callout-note
The interactive versions will not display in this 'live' version of the book. A copy of this page, in the form of a .Rmd document, is available in the Posit Cloud workspace 'Mapping in-class exercise'. You can follow along by opening that and running each cell in turn with the green arrow at the top-right.
:::

```{r}
#| echo: false
#| error: false
#| warning: false
#| message: false

library(tidyverse)
library(plotly)

gapminder_df =  read_tsv('https://raw.githubusercontent.com/jennybc/gapminder/refs/heads/main/data-raw/04_gap-merged.tsv') |> filter(continent!= 'FSU')

titanic_df =  read_csv('https://raw.githubusercontent.com/yann-ryan/infovis-course/master/titanic_df_full.csv')

df <- gapminder_df |>
  filter(year == 2007) |>
  mutate(pop_millions = pop/1000000)
```

## Plotly and ggplotly()

To make interactive plots we can use a library called `plotly.` Plotly has its own syntax for making plots, but it includes a function which will convert any ggplot plot directly into an interactive plotly plot. To use this, simply load the library and add the plot name to the function `ggplotly()`.

On the interactive plot there are a number of things you can do:

-   Hover over a point to get a 'tooltip' displaying more information about that data point.
-   Zoom in, just draw a rectangle on the plot by clicking and dragging the mouse.
-   Toggle any of the categories on or off by clicking the legend points.
-   When you hover over any part of the plot you'll see some display options at the top right.

```{r}
#| error: false
#| warning: false
#| message: false

library(tidyverse)
library(plotly)

p <- df |> 
  ggplot(aes(x = lifeExp, y = gdpPercap, size = pop, fill = continent)) + 
  geom_point()

ggplotly(p)
```

Any of the themes or scales you use will carry over to the interactive plotly version.

**Try it yourself:**

The object `titanic_df` is in your environment. Summarise the data by counting the totals for the `sex` and `embarked` columns, and make a bar chart plotting the sex on the x axis, the total on the y axis, and the embarked location as the fill colour of the bars.

Make the chart interactive using `ggplotly()`.

```{r}

```

## ggplotly() options

ggplotly has some useful options: `dynamicTicks = TRUE` will change the axis ticks as you zoom in. Setting the `tooltip =` argument will allow you to specify the text which is displayed when you hover over a shape with the mouse. By default, the tooltip will display all the aesthetics mapped in your visualisation. In the first chart, for example, we have mapped x, y, size, and fill, and these are all displayed as a label when you hover over a point.

**Try it yourself:**

Copy your bar chart above. Set `dynamicTicks = TRUE` and set the tooltip to display only the `embarked` aesthetic.

```{r}

```

However to get more usefulness out of the interactive it is helpful to be able to make changes to the tooltips, which can be set to any column or series of columns in your data.. Use this to build informative interactives - for example you could even have a small paragraph of text describing something about each point.

Further more, the tooltips can be formatted with HTML, meaning you can construct nicely formatted hover text boxes with custom information.

This is done using `mutate()` and a function called `paste0()`. `paste0()` will concatenate together text, either from columns in your data or from text within quotation marks. Within `paste0()`, separate each piece of text you want to concatenate together with a comma.

To use this new column, map it to the `text` aesthetic, and set it as the tooltip column:

```{r}
p <- df |> 
  mutate(country_continent = paste0("<b>Country: </b>", country, "<br>", "<b>Continent: </b>", continent)) |>
  ggplot(aes(x = lifeExp, y = gdpPercap, text = country_continent, size = pop, fill = continent)) + 
  geom_point()

ggplotly(p, dynamicTicks = TRUE, tooltip = 'text')
```

## Interactive maps with tmap.

### tmap basics

Tmap is also based on the 'grammar of graphics', albeit with its own set of vocabulary.

The basic building block is `tm_shape()` which is the equivalent of `ggplot()`. Further layers are built on top of this, for example `tm_fill()` or `tm_dots()`.

Here are the main layers we can add:

-   [`tm_fill()`](https://r-tmap.github.io/tmap/reference/tm_polygons.html): shaded areas for (multi)polygons

-   [`tm_borders()`](https://r-tmap.github.io/tmap/reference/tm_polygons.html): border outlines for (multi)polygons

-   [`tm_polygons()`](https://r-tmap.github.io/tmap/reference/tm_polygons.html): both, shaded areas and border outlines for (multi)polygons

-   [`tm_lines()`](https://r-tmap.github.io/tmap/reference/tm_lines.html): lines for (multi)linestrings

-   [`tm_symbols()`](https://r-tmap.github.io/tmap/reference/tm_symbols.html): symbols for (multi)points, (multi)linestrings, and (multi)polygons

-   [`tm_raster()`](https://r-tmap.github.io/tmap/reference/tm_raster.html): coloured cells of raster data (there is also [`tm_rgb()`](https://r-tmap.github.io/tmap/reference/tm_rgb.html) for rasters with three layers)

-   [`tm_text()`](https://r-tmap.github.io/tmap/reference/tm_text.html): text information for (multi)points, (multi)linestrings, and (multi)polygons

If we pass an sf object to `tm_shape()` and then add an appropriate layer, we will get a basic map. To map further data to this map, we can map data columns to aesthetics, again, similarly to that we added using ggplot. Here is the most common aesthetics:

-   `fill`: fill colour of a polygon

-   `col`: colour of a polygon border, line, point, or raster

-   `lwd`: line width

-   `lty`: line type

-   `size`: size of a symbol

-   `shape`: shape of a symbol

Set these aesthetics within the relevant shape, e.g. `tm_polygon(fill = 'income_grp')`

To make an interactive map using the worldmap, do the following:

-   Load the tmap library.

-   Download a map from R Natural Earth

-   Switch to the 'view' mode using `tmap_mode("view")` to make the output an interactive map.

-   `tm_shape()` is more or less the equivalent of `ggplot()`: it's the 'base' of the map, where we specify which shapefile should be used. On top of this we add layers, such as polygons or points, and also additional map elements such as a legend, scale, projection, and so forth. We can also add additional `tm_shape()` layers.

-   Add the `tm_polygons()` layer which will draw polygon shapes on the map based on the shapefile. Set the `fill` to the `income_grp` column, in quotation marks. Set the `id` which will determine which column is used to display text with a mouse hover.

-   Add a `tm_view()` layer, which allows us to set options for the interactive viewer. Add a vector of three numbers: the starting latitude and longitude, and the zoom level.

A full worked example of this - including interactive facets with `tm_facets()` - is in the next chapter, [Tmap](51_tmap.qmd#interactive-choropleth-maps).

## In class exercise:

The following is a good example of the kind of real-world workflow in making a map from external data, including the kinds of problems with matching and cleaning you'll often encounter. You'll find it in the previous chapter, [Creating your own geographic objects](47_points-map.qmd#in-class-exercise-2).

## Learning Objectives

Before moving on, take a look and see if you are confident with all of the following learning objectives. If anything is unclear, try to go back over the material, or make a note and ask in our next class!

(the checkboxes are just for your own use, they won't save if you leave or refresh the page)

-   [ ] Convert a ggplot chart into an interactive plotly chart with `ggplotly()`
-   [ ] Customise a plotly tooltip, including with HTML formatting
-   [ ] Know the basic building blocks of tmap (`tm_shape()`, `tm_polygons()`, `tm_symbols()`)
-   [ ] Switch tmap into interactive 'view' mode
