---
title: "Plotly"
---

## If time: interactive charts

Another way to make more readable charts is to include interactive elements. This can help viewers to find interesting points in the data by hovering and clicking, or by zooming in. This negates the need for annotations and labels in many cases, meaning we can make less cluttered visualisations.

These interactive charts will display in the same .html documents we have used for the weekly exercises and assignments.

In R, we can use a library called plotly to easily turn our visualisations interactive.

First, create a regular ggplot plot. Importantly, you should name this plot in your environment by setting a name followed by `=`.

We'll highlight a single country against the rest, and keep the titles and labels, but leave off any extra annotations - plotly's own hover tooltips will do that job for us.

```{r}
#| message: false
#| warning: false

library(tidyverse)

gapminder_df = read_tsv('gapminder_df.csv') |> filter(continent != 'FSU')

europe_df = gapminder_df |> filter(continent == 'Europe')
netherlands_df = gapminder_df |> filter(country == 'Netherlands')

p = ggplot() +
  geom_line(data = europe_df, aes(x = year, y = gdpPercap, group = country), alpha = .5) +
  geom_line(data = netherlands_df, aes(x = year, y = gdpPercap), linewidth = 1.5, color = 'forestgreen') +
  labs(title = "GDP per Capita in Europe",
       subtitle = "1952-2007, Netherlands highlighted",
       caption = "Data from the Gapminder Foundation",
       x = NULL,
       y = "GDP per capita") +
  theme(axis.title = element_text(size = 16))
```

Now, load the library plotly. If it's not installed, install using `install.packages('plotly')`.

You can make plots directly in plotly using this syntax, or you can use a function called `ggplotly()` to automatically convert your static chart. Simply pass your chart name to the function:

```{r}
#| message: false
#| warning: false

library(plotly)

ggplotly(p)
```
